Locally-Oriented Programming:\\ A Simple Programming Model for Stencil-Based Computations on Multi-Level Distributed Memory Architectures

### 2015-01-24 papers/LOPe/coarray-extensions.tex
Initial version.

\section{Programming Model}

The LOPe (for Locally-Oriented Programming extensions) programming model restricts the programmer ---
when implementing a stencil algorithm --- to a local view of the array index space.  Within a LOPe
function, only the local array element (plus a small halo region surrounding the local element) is
visible to the programmer.  This serves to reduce complexity by removing all boundary conditions and
processor topology information from the implementation of the algorithm.

%%This reduction in complexity reduces programming errors.  While developing the convolution example described later in the paper we made an indexing error in applying the 2D stencil loops in the standard serial Fortran test implementation.  This error required over 3 hours of programming time to repair.  First the error had to be isolated to the function implementing the convolution (it was first thought to be in the complicated tiff image output routine as the convolution code was ``thought'' to be too simple to wrong.  Then the index error has to be understood.  As will be seen, LOPe makes it more difficult to make these errors as Fortran array intrics can be used.  In some instance, the restricted semantics of LOPe allows the compiler to catch errors (e.g., some errors involving race conditions).

LOPe was implemented as an embedded domain specific language (DSL).  Fortran was chosen as the base
language for LOPe, although in principle, the same techniques could be applied to language such as C
or C++ (in conjuction with some form of array extensions such as those proposed by Intel).  Fortran
was chosen because it already provides a rich array-based syntax.  Readers who are unfamiliar with
Fortran syntax may wish to consult Appendix A for a brief description of Fortran notation.

\subsection{LOPe Syntax Extensions}

There are only a few syntax additions required for LOPe.  In the code examples describing these
additions (and throughout this paper as well), DSL syntax additions are highlighted by the usage of
capitalization for all new keywords.

\subsubsection{HALO.}

The principle semantic element of LOPe is the concept of a halo region.  A halo is an ``artificial''
or ``virtual'' region surrounding an array that contains boundary-value information.  Halo (also
called ghost-cell) regions are commonly employed to unify array indexing schemes in the vicinity of
an array boundary, so that the array may be referenced using indicies that fall ``outside'' of the
domain of the array.  In LOPe, the halo region is given explicit syntax so that the compiler can
exploit this knowledge.  For example, a halo region can be explicitly provided in a declaration
statement of the form,
\begin{verbatim}
  real, allocatable, dimension(:), HALO(1:*:1) :: A
\end{verbatim}
This statement indicates that array \texttt{A} is one dimensional, will be allocated later when it
is provided with a specific size, and has a halo region of 1 surrounding the array on both sides.
The notation \texttt{M:*:N} is to be read as describing a halo region of size \texttt{M} to the
left, size \texttt{N} to the right, and with an arbitrary size in the middle (denoted by the $*$
symbol).  When used to describe a formal parameter of a function, the \texttt{HALO} keyword may
``assumed'' (from the actual argument by the compiler) as is the following statement,
\begin{verbatim}
  real, HALO(:,:) :: U
\end{verbatim}
which in this instance indicates that \texttt{U} is a two-dimensional array, with a halo region
whose size is ``assumed'' by the compiler from the explicit halo size of the actual array argument
(provided at the calling site of the function).  In LOPe, there is no need in this instance for a
repetitive \texttt{dimension(:,:)} specification as it is inferred by the \texttt{HALO}
specification.

\subsubsection{CONCURRENT.}

The second keyword employed by LOPe (although in this case extended) is \texttt{CONCURRENT}.  The
\texttt{concurrent} keyword already exists in the form of a \texttt{do concurrent} loop construct (a
Fortran 2008 feature).  By using \texttt{do concurrent}, the programmer asserts that specific
iterations of the loop body may be executed by the compiler in \emph{any order;} even
\emph{concurrently.}  For example, the following,
\begin{verbatim}
  pure CONCURRENT subroutine Laplace(U)
     real, HALO(:,:) :: U
     U(0,0) =                 U(0,+1)               &
              +  U(-1,0)  - 3*U(0, 0)  +  U(+1,0)   &
                          +   U(0,-1)
  end subroutine Laplace
\end{verbatim}
defines a function that can be used as part of an iterative solution of Laplace's equation in two
dimensions.  A function with the \texttt{pure} and \texttt{CONCURRENT} attributes may be called from
with a \texttt{do concurrent} loop.  One should imagine that this function is called \emph{for each}
\texttt{(i,j)} iterate of the loop.

\subsubsection{LOPe index notation.}
In the \texttt{Laplace} example shown above, the \texttt{(0,0)} array element is the \emph{local} 
array element and only the local element may be modified.

\input{coarray-comparison}

\subsection{Locally-Oriented Programming}

@@

\begin{verbatim}
   real, allocatable, HALO(-1:*:1) :: A(:)[:], B(:)[:]

   !! Allocate local memory
   !  -------------------------------------------
   allocate(A(N)[*], B(N)[*])
\end{verbatim}

@@

   !! Allocate memory on subimage if available
   !  ----------------------------------------
   if (device /= THIS_IMAGE()) then
      allocate(U1h(0:J)  [device], HALO=(-1:*:1))   [[device]]
      allocate(U2h(0:J/2)[device], HALO=(-1:*:1))   [[device]]
   end if

@@

   do concurrent (j2h = 1:J/2-1)  [[device]]
      call restrict(U1h(2*j2h)[device], U2h(j2h)[device])
   end do

   do concurrent (j2h = 1:J/2-1)  [[device]]
      call restrict(U1h(2*j2h)[device], U2h(j2h)[device], [0.25, 0.5, 0.25])
   end do

   deallocate(U1h, U2h)

   if (device /= THIS_IMAGE()) then
      deallocate(U1h[device], U2h[device])   [[device]]
   end if

>>>>

\section{Coarray Fortran Extensions}

@@

\begin{verbatim}
   real, allocatable, HALO(1:*:1,1:*:1) :: U(:,:)[:]

   !! Allocate normal coarray memory
   !  -------------------------------------------
   allocate(U(M,N)[*])
\end{verbatim}

@@

   !! Allocate memory on subimage if available
   !  ----------------------------------------
   if (device /= THIS_IMAGE()) then
      allocate(U[device], HALO_SRC=U)   [[device]]
   end if

@@

   do i = 1, nLoops
      do concurrent (i=1:M, j=1:N)   [[device]]
         call Laplace(U(i,j)[device])
      end do
      call HALO_TRANSFER(U, BC=CYCLIC)
   end do

   deallocate(U)
   if (device /= THIS_IMAGE()) then
      deallocate(U[device])   [[device]]
   end if
